Staged Abstract Interpreters

### 2019-08-15 paper/10_intro.tex
typos

In recent years, we observe rich progress on the methodologies for
constructing abstract interpreters from systematic principles rather than
ad-hoc engineering. Of particular note is the Abstracting Abstract
Machine (AAM) methodology \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10}. This methodology uncovers an approach to derive sound
abstract interpreters from their concrete counterparts, for example,
abstract machines, where the soundness can be easily established by
examining the transformation of semantic artifacts. For example, the
CEK machine \cite{DBLP:conf/popl/FelleisenF87} for concrete execution
can be readily refactored to an effective $0$-CFA control-flow
analysis \cite{Shivers:1988:CFA:53990.54007,
  Midtgaard:2012:CAF:2187671.2187672}: first, tweak the environment
dereference as a nondeterministic choice, such that the environment
may contain mutiple possible values for a varaiable, then allocate
continuations in the environment to exploit return-flow information,
and also constrain the addresses space to be finite.  This systematic
abstraction approach can be tailored to different language features
(such as states, first-class controls, exceptions, and concurrency) and
sensitivity analyses \cite{DBLP:conf/icfp/Gilray0M16,
  DBLP:conf/popl/GilrayL0MH16, Darais:2015:GTM:2814270.2814308}. It
has also been applied to various small-step abstract machines
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10,
  Sergey:2013:MAI:2491956.2491979} and big-step definitional
interpreters \cite{Wei:2018:RAA:3243631.3236800,
  DBLP:journals/pacmpl/DaraisLNH17, Keidel:2018:CSP:3243631.3236767}.

@@

However, besides the intrinsic complexity of static analysis, there
are additional abstraction penalties with these high-level
implementation approaches.  First, similar to concrete interpreters,
the abstract interpreter analyzes the program by traversing the
abstract syntax tree (AST), which poses an interpretive overhead, such as
the pattern matching on the ASTs and recursive calls on the subexpressions.
Such overhead can be negligible if the abstract
interpreter only runs on the program a single time, but can also
accumulate significantly if run repeatedly (for example, on
libraries).  Second, the abstract interpreter written in pure languages
usually extensively uses effect systems to implement the semantics of
abstract interpretation.  For example, the abstract interpreter that
returns all possible runtime values is a form of nondeterminism, where
nondeterminism monads can help.  Although such pure approaches have
their own merits and elegance, compared with imperative stateful
implementations, they are significantly slower.

@@

Partial evaluation \cite{DBLP:books/daglib/0072559} was the first
proposed approach to realize Futamura projections: it first discovers
the binding-times (\textit{static} or \textit{dynamic}) of variables,
then evaluates the static part of the program, and generates a
residualized program that solely relies on the dynamic part. However,
it is hard to precisely analyze binding-times given an arbitrary
program.  As an alternative approach to specialization, multi-stage
programming (MSP) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} lets
the programmer to explicitly control and annotate the binding-times in
the program, then the MSP system will check whether these annotations
are consistent and specialize the program using that information.  The
staging annotations can be either syntactic (e.g., quote and
quasiquote in MetaML/MetaOCaml) or type-based (e.g., the Lightweight
Modular Staging framework \cite{DBLP:conf/gpce/RompfO10} in Scala).

@@

Program specialization and abstract interpretation are two orthogonal
concepts.  To implement the confluence of them, we first construct a
generic interpreter that is agnostic to both binding-times and value
domains used in the semantics.  Later, the generic interpreter can be
instantiated from these two dimensions (Figure~\ref{confluence}):
\begin{itemize}
\item With a flat binding-time and concrete domains, it is a ordinary
  definitional interpreter based on big-step operational semantics;
\item With two-level binding-times and concrete domain, it is a
  compiler that translate to program intro antoher language;
\item With a flat binding-time and abstract domains, it is an
  definitional abstract interpreter \cite{DBLP:journals/pacmpl/DaraisLNH17}
  that statically computes runtime properties;
\item With two-level binding-times and abstract domains, it is an optimizing
  program analyzer but works in the fashion of compilation.
\end{itemize}

@@

\paragraph{Contributions} Briefly, the contribution of the paper is as follows:
\begin{itemize}[leftmargin=2em]
  \item To our best knowledge, we present the first one to use
    multi-stage programming to specialize a general abstract interpreter,
    and our approach does not touch the soundness of analyses.
  \item Intellectually, our framework naturally extends the first
    Futamura projection to abstract interpreters, showing a
    well-grounded approach to optimize static analyses via
    metaprogramming.
  \item Practically, we show that staging an abstract interpreter is
    useful to improve performance and scalability of analyses by case
    studies and an empirical evaluation.
\end{itemize}

@@

\iffalse
On the other side, static analysis is a tradeoff between performance and
precision: higher precision usually leads to longer running time.

>>>>

In recent years, we observe rich progress on the methodologies for
constructing abstract interpreters from systematic principles rather than
ad-hoc engineering. Of particular note is the Abstracting Abstract
Machine (AAM) methodology \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10}. This methodology uncovers an approach to derive sound
abstract interpreters from their concrete counterparts, for example,
abstract machines, where the soundness can be easily established by
examining the transformation of semantic artifacts. For example, the
CEK machine \cite{DBLP:conf/popl/FelleisenF87} for concrete execution
can be readily refactored to an effective $0$-CFA control-flow
analysis \cite{Shivers:1988:CFA:53990.54007,
  Midtgaard:2012:CAF:2187671.2187672}: first, tweak the environment
dereference as a nondeterministic choice, such that the environment
may contain multiple possible values for a variable, then allocate
continuations in the environment to exploit return-flow information,
and also constrain the addresses space to be finite.  This systematic
abstraction approach can be tailored to different language features
(such as states, first-class controls, exceptions, and concurrency) and
sensitivity analyses \cite{DBLP:conf/icfp/Gilray0M16,
  DBLP:conf/popl/GilrayL0MH16, Darais:2015:GTM:2814270.2814308}. It
has also been applied to various small-step abstract machines
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10,
  Sergey:2013:MAI:2491956.2491979} and big-step definitional
interpreters \cite{Wei:2018:RAA:3243631.3236800,
  DBLP:journals/pacmpl/DaraisLNH17, Keidel:2018:CSP:3243631.3236767}.

@@

However, besides the intrinsic complexity of static analysis, there
are additional abstraction penalties with these high-level
implementation approaches.  First, similar to concrete interpreters,
the abstract interpreter analyzes the program by traversing the
abstract syntax tree (AST), which poses an interpretive overhead, such as
the pattern matching on the ASTs and recursive calls on the sub-expressions.
Such overhead can be negligible if the abstract
interpreter only runs on the program a single time, but can also
accumulate significantly if run repeatedly (for example, on
libraries).  Second, the abstract interpreter written in pure languages
usually extensively uses effect systems to implement the semantics of
abstract interpretation.  For example, the abstract interpreter that
returns all possible runtime values is a form of nondeterminism, where
nondeterminism monads can help.  Although such pure approaches have
their own merits and elegance, compared with imperative stateful
implementations, they are significantly slower.

@@

Partial evaluation \cite{DBLP:books/daglib/0072559} was the first
proposed approach to realize Futamura projections: it first discovers
the binding-times (\textit{static} or \textit{dynamic}) of variables,
then evaluates the static part of the program, and generates a
residualized program that solely relies on the dynamic part. However,
it is hard to precisely analyze binding-times given an arbitrary
program.  As an alternative approach to specialization, multi-stage
programming (MSP) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} lets
the programmer to explicitly control and annotate the binding-times in
the program, then the MSP system will check whether these annotations
are consistent and specialize the program using that information.  The
staging annotations can be either syntactic (e.g., quote and
quasi-quote in MetaML/MetaOCaml) or type-based (e.g., the Lightweight
Modular Staging framework \cite{DBLP:conf/gpce/RompfO10} in Scala).

@@

Program specialization and abstract interpretation are two orthogonal
concepts.  To implement the confluence of them, we first construct a
generic interpreter that is agnostic to both binding-times and value
domains used in the semantics.  Later, the generic interpreter can be
instantiated from these two dimensions (Figure~\ref{confluence}):
\begin{itemize}
\item With a flat binding-time and concrete domains, it is a ordinary
  definitional interpreter based on big-step operational semantics;
\item With two-level binding-times and concrete domain, it is a
  compiler that translate to program intro another language;
\item With a flat binding-time and abstract domains, it is an
  definitional abstract interpreter \cite{DBLP:journals/pacmpl/DaraisLNH17}
  that statically computes runtime properties;
\item With two-level binding-times and abstract domains, it is an optimizing
  program analyzer but works in the fashion of compilation.
\end{itemize}

@@

\paragraph{Contributions} Briefly, the contribution of the paper is as follows:
\begin{itemize}[leftmargin=2em]
  \item To our best knowledge, we present the first one to use
    multi-stage programming to specialize a general abstract interpreter,
    and our approach does not touch the soundness of analyses.
  \item Intellectually, our framework naturally extends the first
    Futamura projection to abstract interpreters, showing a
    well-grounded approach to optimize static analyses via
    meta-programming.
  \item Practically, we show that staging an abstract interpreter is
    useful to improve performance and scalability of analyses by case
    studies and an empirical evaluation.
\end{itemize}

@@

\iffalse
On the other side, static analysis is a trade-off between performance and
precision: higher precision usually leads to longer running time.

### 2019-08-15 paper/20_prelim.tex
typos

Monad transformers are type constructors of kind @(* -> *) -> (* -> *)@, which
take a monad as argument and produces another monad. By using monad
transformers, we can combine multiple monads into a single one. Constructing
extensible interpreters using monad transformers was first explored by
\citet{DBLP:conf/popl/LiangHJ95}, and later has been applied to abstract
interpreters \cite{Sergey:2013:MAI:2491956.2491979,
  DBLP:journals/pacmpl/DaraisLNH17, Darais:2015:GTM:2814270.2814308}.
In this section, we present the generic inferface in the style of big-step
definitional interpreter. The key idea is to keep the binding-time type and
returned monadic type abstract, so that they can be instantiated differently.
We also need to abstract the primitive operations on those types.

\paragraph{Basic Types} We start with some basic type definitions used in the
interpreter. The identifiers in the program are represented by
strings. To represent states, two required components for the
interpreter are environments @Env@ and stores @Store@. @Env@ maps from
identifiers to addresses, and @Store@ maps from addresses to values,
respectively. The @Env@ captures bound variables in scope, and @Store@
models a persistent heap through the program runtime.  At this moment,
the domains of addresses and values are still abstract, i.e., they are
declared as abstract types.
\begin{lstlisting}
  trait Semantics {
    type Ident = String; type Addr; type Value
    type Env = Map[Ident, Addr]; type Store = Map[Addr, Value]
    type R[_] // Binding-time as a higher-kinded type
    ... // The definitions in the rest of this section are enclosed in trait Semantics.
  }
\end{lstlisting}

\paragraph{Binding-time Abstraction} As mentioned before, the binding-time is
declared as a higher-kinded type @R[_]@. We are free to use @R@ to annotate
other datatypes in the interpreter; it is also injected into the monadic
type @MonadOps@.
If we simply instantiate @R@ an identity type (i.e., @type R[T] = T@),
then the generic interpreter is a standard definitional interpreter
that will execute the program.  In Section \ref{stagedinterp}, we will
instantiate @R@ using LMS's built-in next-stage type annotation @Rep@,
which makes the interpreter act as a compiler.

@@

We also declare several operations to mannipulate environments and
stores. These methods return monadic values of type @AnsM[_]@, which
may encapsulate over the environment or store, or simply a @Unit@
value for effects. For examples, @local_env@ installs a new environment
@ρ@ when evaluating the monadic value @ans@; @set_store@ takes a pair of
addresses and values (potentially staged) and updates the store accordingly.

@@

\paragraph{The Interpreter} Now we construct the semantics-agnostic interpreter
in monadic form, shown in Figure \ref{fig:shared_int}. The basic idea
of a generic interpretation is to traverse the abstract syntax tree
while maintaining the effects such as reader and state.  Note that the
interpreter is written in open-recursive style -- it can not refer to
itself directly, instead, @eval@ takes an additional parameter @ev@ of
type @Expr => Ans@ that refers to itself. Consequently, the method
@close@ that lifts lambda terms to closures and method @ap_clo@ that
applies functions also takes an extra @ev@, because evaluation may
happen inside of them.
To close the open-recursion, we use a fixed-point opertor @fix@.
For concrete-interpretation instantiation, it works like the Y
combinator; for abstract-interpretation instantiation, it instruments
the interpreter by memoizing its input and output, which will ensure
the termination of abstract interpretation.
Finally, a top-level wrapper @run@ is declared; the return type
@Result@ depends on what kind of monad we will be using, this is also
an abstract type.
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans
  type Result; def run(e: Expr): Result
\end{lstlisting}

@@

Here we sketch the basic idea of @ReaderT@ and @StateT@. Readers may
refer to \cite{DBLP:conf/popl/LiangHJ95, Chiusano:2014:FPS:2688794} for more detail.
A @ReaderT@ monad transformer encapsulates computation @R => M[A]@, where
@R@ is the reader type, and @M[_]@ is the inner monad type.
Given a @R@, @ReaderT@ monad produecs a transformed value of type @M[A]@.
Similarly, a @StateT@ monad encapsulates computation @S => M[(A, S)]@, where
@S@ is the state type, and @M[_]@ is the inner monad type.
Given a @S@, @StateT@ monad produces a transformed value of type @M[(A, S)]@,
where the new state (type @S@) is accompanied with the value (type @A@).
Note that for the moment, the binding-time type @R@ is an identity type, thus
these monads operate on unstaged data. We can also see this from the
signature of @flatMap@: the argument function @f@ takes an unstaged value of type @A@ and
produces a monadic value. In the following code, we elide operations other than @flatMap@.
\begin{lstlisting}
  case class ReaderT[M[_]: Monad, R, A](run: R => M[A]) {
    def flatMap[B](f: A => ReaderT[M, R, B]): ReaderT[M, R, B] =
      ReaderT(r => Monad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class StateT[M[_]: Monad, S, A](run: S => M[(A, S)]) {
    def flatMap[B](f: A => StateT[M, S, B]): StateT[M, S, B] =
      StateT(s => Monad[M].flatMap(run(s)) { case (a, s1) => f(a).run(s1) }); ... }
\end{lstlisting}

>>>>

Monad transformers are type constructors of kind @(* -> *) -> (* -> *)@, which
take a monad as argument and produces another monad. By using monad
transformers, we can combine multiple monads into a single one. Constructing
extensible interpreters using monad transformers was first explored by
\citet{DBLP:conf/popl/LiangHJ95}, and later has been applied to abstract
interpreters \cite{Sergey:2013:MAI:2491956.2491979,
  DBLP:journals/pacmpl/DaraisLNH17, Darais:2015:GTM:2814270.2814308}.
In this section, we present the generic interface in the style of big-step
definitional interpreter. The key idea is to keep the binding-time type and
returned monadic type abstract, so that they can be instantiated differently.
We also need to abstract the primitive operations on those types.

\paragraph{Basic Types} We start with some basic type definitions used in the
interpreter. The identifiers in the program are represented by
strings. To represent states, two required components for the
interpreter are environments @Env@ and stores @Store@. @Env@ maps from
identifiers to addresses, and @Store@ maps from addresses to values,
respectively. The @Env@ captures bound variables in scope, and @Store@
models a persistent heap through the program run-time.  At this moment,
the domains of addresses and values are still abstract, i.e., they are
declared as abstract types.
\begin{lstlisting}
  trait Semantics {
    type Ident = String; type Addr; type Value
    type Env = Map[Ident, Addr]; type Store = Map[Addr, Value]
    type R[_] // Binding-time as a higher-kinded type
    ... // The definitions in the rest of this section are enclosed in trait Semantics.
  }
\end{lstlisting}

\paragraph{Binding-time Abstraction} As mentioned before, the binding-time is
declared as a higher-kinded type @R[_]@. We are free to use @R@ to annotate
other data-types in the interpreter; it is also injected into the monadic
type @MonadOps@.
If we simply instantiate @R@ an identity type (i.e., @type R[T] = T@),
then the generic interpreter is a standard definitional interpreter
that will execute the program.  In Section \ref{stagedinterp}, we will
instantiate @R@ using LMS's built-in next-stage type annotation @Rep@,
which makes the interpreter act as a compiler.

@@

We also declare several operations to manipulate environments and
stores. These methods return monadic values of type @AnsM[_]@, which
may encapsulate over the environment or store, or simply a @Unit@
value for effects. For examples, @local_env@ installs a new environment
@ρ@ when evaluating the monadic value @ans@; @set_store@ takes a pair of
addresses and values (potentially staged) and updates the store accordingly.

@@

\paragraph{The Interpreter} Now we construct the semantics-agnostic interpreter
in monadic form, shown in Figure \ref{fig:shared_int}. The basic idea
of a generic interpretation is to traverse the abstract syntax tree
while maintaining the effects such as reader and state.  Note that the
interpreter is written in open-recursive style -- it can not refer to
itself directly, instead, @eval@ takes an additional parameter @ev@ of
type @Expr => Ans@ that refers to itself. Consequently, the method
@close@ that lifts lambda terms to closures and method @ap_clo@ that
applies functions also takes an extra @ev@, because evaluation may
happen inside of them.
To close the open-recursion, we use a fixed-point operator @fix@.
For concrete-interpretation instantiation, it works like the Y
combinator; for abstract-interpretation instantiation, it instruments
the interpreter by memoizing its input and output, which will ensure
the termination of abstract interpretation.
Finally, a top-level wrapper @run@ is declared; the return type
@Result@ depends on what kind of monad we will be using, this is also
an abstract type.
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans
  type Result; def run(e: Expr): Result
\end{lstlisting}

@@

Here we sketch the basic idea of @ReaderT@ and @StateT@. Readers may
refer to \cite{DBLP:conf/popl/LiangHJ95, Chiusano:2014:FPS:2688794} for more detail.
A @ReaderT@ monad transformer encapsulates computation @R => M[A]@, where
@R@ is the reader type, and @M[_]@ is the inner monad type.
Given a @R@, @ReaderT@ monad produces a transformed value of type @M[A]@.
Similarly, a @StateT@ monad encapsulates computation @S => M[(A, S)]@, where
@S@ is the state type, and @M[_]@ is the inner monad type.
Given a @S@, @StateT@ monad produces a transformed value of type @M[(A, S)]@,
where the new state (type @S@) is accompanied with the value (type @A@).
Note that for the moment, the binding-time type @R@ is an identity type, thus
these monads operate on unstaged data. We can also see this from the
signature of @flatMap@: the argument function @f@ takes an unstaged value of type @A@ and
produces a monadic value. In the following code, we elide operations other than @flatMap@.
\begin{lstlisting}
  case class ReaderT[M[_]: Monad, R, A](run: R => M[A]) {
    def flatMap[B](f: A => ReaderT[M, R, B]): ReaderT[M, R, B] =
      ReaderT(r => Monad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class StateT[M[_]: Monad, S, A](run: S => M[(A, S)]) {
    def flatMap[B](f: A => StateT[M, S, B]): StateT[M, S, B] =
      StateT(s => Monad[M].flatMap(run(s)) { case (a, s1) => f(a).run(s1) }); ... }
\end{lstlisting}

### 2019-08-15 paper/30_stageconc.tex
Titles and spaces
typos

Lightweight modular staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a
multi-stage programming framework implemented as a Scala library that enables
dynamic code generation in a type-safe manner. Different from the approach of
MetaML/MetaOCaml \cite{DBLP:conf/flops/Kiselyov14, DBLP:conf/gpce/CalcagnoTHL03} that
uses syntactic quotations and quasiquotations, LMS distinguishes binding-time
based on types. LMS provides a type constructor @Rep[T]@ where @T@ can be
an arbitrary type, indicating a value of @T@ will be known at the next stage.
All operations acting on @Rep[T]@ expressions will be residualized as generated
code.

@@

The LMS framework provides staging support for primitive data types such as
@Int@ and @Double@, and commonly-used data structures such as lists and maps.
The idea behind the framework is to construct a sea-of-node intermediate
representations (IR) for the next stage program at the current stage
\cite{DBLP:conf/birthday/Rompf16}. For convinence, the conversion from
expressions of type @Rep[_]@ to their IR is done by using the implicit design
pattern. As we will see later, implementing staging and code generation
support for user-defined classes is also straightforward.

@@

\paragraph{Staged Monads}
We use the same structure of monad stack in the unstaged interpreter: a reader
monad with a state monad. But now the monads operate on staged values, for
brivity, we call them \textit{staged monads}. In the code, we also use a @Rep@
prefix on the constructors and types to differentiate them. But it is important
to note that objects of a case class @ReaderT@/@StateT@ are not staged, the
monadic computation like @R => M[A]@ are also not staged. Instead, the internal
data that these monads operate on are staged, i.e., @R@ and @A@ in the
reader monad, @S@ and @A@ in the state monad. The following code
snippet shows the idea. We use \hl{light gray} to highlight where @Rep@ type is added:
\begin{lstlisting}[escapechar=!]
  case class RepReaderT[M[_]: RepMonad, R, A](run: !\hl{Rep[R]}! => M[A]) {
    def flatMap[B](f: !\hl{Rep[A]}! => ReaderT[M, R, B]): RepReaderT[M, R, B] =
      RepReaderT(r => RepMonad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class RepStateT[M[_]: RepMonad, S, A](run: !\hl{Rep[S]}! => M[(A, S)]) {
    def flatMap[B](f: !\hl{Rep[A]}! => StateT[M, S, B]): RepStateT[M, S, B] =
      RepStateT(s => RepMonad[M].flatMap(run(s))(as => f(as._1).run(as._2))); ... }
\end{lstlisting}

@@

Readers may notice that the conversion between unstaged monads and staged monads
is merely changing the type of unstaged data to staged data, which in fact can
be achieved without modifying the implementation of monads. This is true so far,
but as we will see, it is not straightfoward when for nondeterminism monad
(@SetT@) in abstract interpreters. Because not only the elements in the set are
staged, but the whole set is also staged, and we have no knowledge about how
many elements in the set. In this section, we explicitly distinguish the two
versions of monadic interfaces, later we will instantiated staged set monad
using a different strategy.

\paragraph{Primitive Operations} Most of the primitive operations can be easily
translated to their staged versions -- we just need to change the types.
As we mentioned before, we will take more care on how functions and
applications are handled. The problem here is what should we do for
lambda terms when staging? We cannot create a next-stage
defunctionalized closure for it, because it means we still need to
access the interpreter at the next stage. The desired goal is to eliminate the
interpretation overhead staging, so the right thing is compiling the lambda term
and its environment to a next-stage Scala function. To do this, we need to
recursively call the interpreter function, i.e., the @ev@ argument passed to @close@.
The following code implements the idea:

\begin{lstlisting}
  type ValSt = (Value, Store)
  def emit_compiled_clo(f: (Rep[Value], Rep[Store]) => Rep[ValSt], λ: Lam, ρ: Exp[Env]): Rep[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value], Rep[Store]) => Rep[ValSt] = { case (v: Rep[Value], σ: Rep[Store]) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ + (α → v))
    }; emit_compiled_clo(f, λ, ρ)
  }
\end{lstlisting}

We have created a current-stage function @f@ that takes a next-stage value and
store, and returns staged value. Then we use @emit_complied_clo@ to delay @f@ to
a next-stage Scala function, represented by the type @Rep[Value]@. Inside of
@f@, we can access the evaluator via @ev@, but @ev@ produces a monadic value of
type @AnsM@, which can only exist at the current stage. To connect the
current-stage monadic values and future-stage grounded values, the evaluator
@ev@ is invoked on the body expression @e@ to be specialized w.r.t. @e@, and
then we \textit{collpase} the @AnsM@ monad to grounded values by providing the
prepared environment and store.

For function applications @ap_clo@, we have two \textit{next-stage} values @fun@
and @arg@, but what we can do on these values of @Rep@ type? In fact, @fun@ does
not have an intensional application operation we can use directly; but we know
that it is a staged value produced by the @close@ function, which represents a
next-stage Scala function. Therefore we can just generate a next-stage function
application, also in Scala. Since @ap_clo@ returns a monadic value of type
@Ans@, we can still use the monadic operation @get_store@ to obtain the latest
store object. Then we use @emit_ap_clo@ to generate a current-stage
representation of the future-stage application result, which is a pair of values
and stores @ValSt@. After getting that result @vs@, we use @put_store@ to
replace the store and reify the result to the current-stage monadic value.

\begin{lstlisting}
  def emit_ap_clo(fun: Rep[Value], arg: Rep[Value], σ: Rep[Store]): Rep[ValSt]
  def ap_clo(ev: EvalFun)(fun: Rep[Value], arg: Rep[Value]): Ans = for {
    σ  <- get_store
    vs <- lift[ValSt](emit_ap_clo(fun, arg, σ))
    _  <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

Compared with the unstaged version, we can observe a shift of using the
evaluation function @ev@ from application-time to value-representation-time: in
the unstaged interpreter, the invoke of @ev@ happens in @ap_clo@; while in the
staged interpreter, we eagerly apply @ev@ when denotating the lambda term to our
value domain.

\subsection{A Little Bit Code Generation}

In the previous section, we merely show the type of @emit_compiled_clo@ and
@emit_ap_clo@ without a concrete implemtation. In this section, we brifely
discuss the IR node generated by them and sketch code generation.
We define the IR nodes @IRCompiledClo@ and @IRApClo@ using @case class@
extending from @Def[T]@. @Def[T]@ is a built-in type in the LMS framework,
representing next-stage definitions of type @T@.

\begin{lstlisting}
  case class IRCompiledClo(f: Rep[(ValSt) => ValSt], λ: Lam, ρ: Rep[Env]) extends Def[Value]
  case class IRApClo(rator: Rep[Value], rand: Rep[Value], σ: Rep[Store]) extends Def[ValSt]
\end{lstlisting}

The IR nodes are manipulated by the LMS passes and finally generated to
next-stage Scala code. To generate code, LMS provides an @emitNode@ method for
programmers to control what is generated for each kind of node. Here we have
matched @IRCompiledClo@ and @IRApClo@ and will generate definitions for them.
For compiled closures, we put the function @f@ and its companying syntactic term
and environment to its next-stage value representation @CompiledClo@. For
applications, we have the knowledge that @fun@ is a @CompiledClo@ at the stage,
and it has a callable field @f@, so we generate a next-stage function
application @f(arg, σ)@. The @quote@s are invocations to generate code for the
other next-stage values.

\begin{lstlisting}
override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
  case IRCompiledClo(f, λ, ρ) => emitValDef(sym, s"CompiledClo($\textdollar${quote(f)}, $\textdollar${quote(λ)}, $\textdollar${quote(ρ)})")
  case IRApClo(rator, rand, σ) => emitValDef(sym, s"$\textdollar$rator.f($\textdollar${quote(rand)}, $\textdollar${quote(σ)})")
  ...
}
\end{lstlisting}

>>>>

Lightweight modular staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a
multi-stage programming framework implemented as a Scala library that enables
dynamic code generation in a type-safe manner. Different from the approach of
MetaML/MetaOCaml \cite{DBLP:conf/flops/Kiselyov14, DBLP:conf/gpce/CalcagnoTHL03} that
uses syntactic quotations and quasi-quotations, LMS distinguishes binding-time
based on types. LMS provides a type constructor @Rep[T]@ where @T@ can be
an arbitrary type, indicating a value of @T@ will be known at the next stage.
All operations acting on @Rep[T]@ expressions will be residualized as generated
code.

@@

The LMS framework provides staging support for primitive data types such as
@Int@ and @Double@, and commonly-used data structures such as lists and maps.
The idea behind the framework is to construct a sea-of-node intermediate
representations (IR) for the next stage program at the current stage
\cite{DBLP:conf/birthday/Rompf16}. For convenience, the conversion from
expressions of type @Rep[_]@ to their IR is done by using the implicit design
pattern. As we will see later, implementing staging and code generation
support for user-defined classes is also straightforward.

@@

\paragraph{Staged Monads}
We use the same structure of monad stack in the unstaged interpreter: a reader
monad with a state monad. But now the monads operate on staged values, for
brevity, we call them \textit{staged monads}. In the code, we also use a @Rep@
prefix on the constructors and types to differentiate them. But it is important
to note that objects of a case class @ReaderT@/@StateT@ are not staged, the
monadic computation like @R => M[A]@ are also not staged. Instead, the internal
data that these monads operate on are staged, i.e., @R@ and @A@ in the
reader monad, @S@ and @A@ in the state monad. The following code
snippet shows the idea. We use \hl{light gray} to highlight where @Rep@ type is added:
\begin{lstlisting}[escapechar=!]
  case class RepReaderT[M[_]: RepMonad, R, A](run: !\hl{Rep[R]}! => M[A]) {
    def flatMap[B](f: !\hl{Rep[A]}! => ReaderT[M, R, B]): RepReaderT[M, R, B] =
      RepReaderT(r => RepMonad[M].flatMap(run(r))(a => f(a).run(r))); ... }
  case class RepStateT[M[_]: RepMonad, S, A](run: !\hl{Rep[S]}! => M[(A, S)]) {
    def flatMap[B](f: !\hl{Rep[A]}! => StateT[M, S, B]): RepStateT[M, S, B] =
      RepStateT(s => RepMonad[M].flatMap(run(s))(as => f(as._1).run(as._2))); ... }
\end{lstlisting}

@@

Readers may notice that the conversion between unstaged monads and staged monads
is merely changing the type of unstaged data to staged data, which in fact can
be achieved without modifying the implementation of monads. This is true so far,
but as we will see, it is not straightforward when for nondeterminism monad
(@SetT@) in abstract interpreters. Because not only the elements in the set are
staged, but the whole set is also staged, and we have no knowledge about how
many elements in the set. In this section, we explicitly distinguish the two
versions of monadic interfaces, later we will instantiated staged set monad
using a different strategy.

\paragraph{Primitive Operations} Most of the primitive operations can be easily
translated to their staged versions -- we just need to change the types.
As we mentioned before, we will take more care on how functions and
applications are handled. The problem here is what should we do for
lambda terms when staging? We cannot create a next-stage
defunctionalized closure for it, because it means we still need to
access the interpreter at the next stage. The desired goal is to eliminate the
interpretation overhead staging, so the right thing is compiling the lambda term
and its environment to a next-stage Scala function. To do this, we need to
recursively call the interpreter function, i.e., the @ev@ argument passed to @close@.
The following code implements the idea:
\begin{lstlisting}
  type ValSt = (Value, Store)
  def emit_compiled_clo(f: (Rep[Value], Rep[Store]) => Rep[ValSt], λ: Lam, ρ: Exp[Env]): Rep[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value], Rep[Store]) => Rep[ValSt] = { case (v: Rep[Value], σ: Rep[Store]) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ + (α → v))
    }; emit_compiled_clo(f, λ, ρ)
  }
\end{lstlisting}

We have created a current-stage function @f@ that takes a next-stage value and
store, and returns staged value. Then we use @emit_complied_clo@ to delay @f@ to
a next-stage Scala function, represented by the type @Rep[Value]@. Inside of
@f@, we can access the evaluator via @ev@, but @ev@ produces a monadic value of
type @AnsM@, which can only exist at the current stage. To connect the
current-stage monadic values and future-stage grounded values, the evaluator
@ev@ is invoked on the body expression @e@ to be specialized w.r.t. @e@, and
then we \textit{collapse} the @AnsM@ monad to grounded values by providing the
prepared environment and store.

For function applications @ap_clo@, we have two \textit{next-stage} values @fun@
and @arg@, but what we can do on these values of @Rep@ type? In fact, @fun@ does
not have an intensional application operation we can use directly; but we know
that it is a staged value produced by the @close@ function, which represents a
next-stage Scala function. Therefore we can just generate a next-stage function
application, also in Scala. Since @ap_clo@ returns a monadic value of type
@Ans@, we can still use the monadic operation @get_store@ to obtain the latest
store object. Then we use @emit_ap_clo@ to generate a current-stage
representation of the future-stage application result, which is a pair of values
and stores @ValSt@. After getting that result @vs@, we use @put_store@ to
replace the store and reify the result to the current-stage monadic value.
\begin{lstlisting}
  def emit_ap_clo(fun: Rep[Value], arg: Rep[Value], σ: Rep[Store]): Rep[ValSt]
  def ap_clo(ev: EvalFun)(fun: Rep[Value], arg: Rep[Value]): Ans = for {
    σ  <- get_store
    vs <- lift[ValSt](emit_ap_clo(fun, arg, σ))
    _  <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

Compared with the unstaged version, we can observe a shift of using the
evaluation function @ev@ from application-time to value-representation-time: in
the unstaged interpreter, the invoke of @ev@ happens in @ap_clo@; while in the
staged interpreter, we eagerly apply @ev@ when denoting the lambda term to our
value domain.

\subsection{A Little Bit of Code Generation}

In the previous section, we merely show the type of @emit_compiled_clo@ and
@emit_ap_clo@ without a concrete implementation. In this section, we briefly
discuss the IR node generated by them and sketch code generation.
We define the IR nodes @IRCompiledClo@ and @IRApClo@ using @case class@
extending from @Def[T]@. @Def[T]@ is a built-in type in the LMS framework,
representing next-stage definitions of type @T@.
\begin{lstlisting}
  case class IRCompiledClo(f: Rep[(ValSt) => ValSt], λ: Lam, ρ: Rep[Env]) extends Def[Value]
  case class IRApClo(rator: Rep[Value], rand: Rep[Value], σ: Rep[Store]) extends Def[ValSt]
\end{lstlisting}

The IR nodes are manipulated by the LMS passes and finally generated to
next-stage Scala code. To generate code, LMS provides an @emitNode@ method for
programmers to control what is generated for each kind of node. Here we have
matched @IRCompiledClo@ and @IRApClo@ and will generate definitions for them.
For compiled closures, we put the function @f@ and its accompanying syntactic term
and environment to its next-stage value representation @CompiledClo@. For
applications, we have the knowledge that @fun@ is a @CompiledClo@ at the stage,
and it has a callable field @f@, so we generate a next-stage function
application @f(arg, σ)@. The @quote@s are invocations to generate code for the
other next-stage values.
\begin{lstlisting}
override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
  case IRCompiledClo(f, λ, ρ) => emitValDef(sym, s"CompiledClo($\textdollar${quote(f)}, $\textdollar${quote(λ)}, $\textdollar${quote(ρ)})")
  case IRApClo(rator, rand, σ) => emitValDef(sym, s"$\textdollar$rator.f($\textdollar${quote(rand)}, $\textdollar${quote(σ)})")
  ...
}
\end{lstlisting}
